---
title: "ESP_CTD_PCA"
author: "Markus Min"
date: "1/28/2021"
output: html_document
---

### Dependencies

In addition to the R packages required for this script, you will also need to install **qiime2** and **DEICODE** (a qiime2 plugin) and create a qiime2 environment to run DEICODE within.


### Load libraries
```{r}
library(tidyverse)
library(phyloseq)
library(qiime2R)
library(here)
```

### Load data

```{r}
asv_table_path_COI <- here::here("data", "COI", "ESP_CTD_COI_asv_table.csv")
tax_table_path_COI <- here::here("data", "COI", "ESP_CTD_COI_tax_table.csv")
metadata_path_COI <- here::here("data", "COI", "ESP_CTD_COI_metadata.csv")

ESP_CTD_COI_phyloseq <- merge_phyloseq(otu_table(read.csv(asv_table_path_COI, row.names = 1),taxa_are_rows = TRUE),
                                       tax_table(as.matrix(read.csv(tax_table_path_COI, row.names = 1))),
                                       sample_data(read.csv(metadata_path_COI, row.names = 1)))

#### 18S
asv_table_path_18S <- here::here("data", "18S", "ESP_CTD_18S_asv_table.csv")
tax_table_path_18S <- here::here("data", "18S", "ESP_CTD_18S_tax_table.csv")
metadata_path_18S <- here::here("data", "18S", "ESP_CTD_18S_metadata.csv")

ESP_CTD_18S_phyloseq <- merge_phyloseq(otu_table(read.csv(asv_table_path_18S, row.names = 1),taxa_are_rows = TRUE),
                                       tax_table(as.matrix(read.csv(tax_table_path_18S, row.names = 1))),
                                       sample_data(read.csv(metadata_path_18S, row.names = 1)))

#### 12S
asv_table_path_12S <- here::here("data", "12S", "ESP_CTD_12S_asv_table.csv")
tax_table_path_12S <- here::here("data", "12S", "ESP_CTD_12S_tax_table.csv")
metadata_path_12S <- here::here("data", "12S", "ESP_CTD_12S_metadata.csv")

ESP_CTD_12S_phyloseq <- merge_phyloseq(otu_table(read.csv(asv_table_path_12S, row.names = 1),taxa_are_rows = TRUE),
                                       tax_table(as.matrix(read.csv(tax_table_path_12S, row.names = 1))),
                                       sample_data(read.csv(metadata_path_12S, row.names = 1)))




# Set figure directory
fig_dir <- here::here("figures", "PCAs")
```

# COI

### Prep data for DEICODE
```{r}
# Set directory for saving DEICODE data
deicode_dir <- here::here("18S_12S_CO1_Analyses", "DEICODE", "COI")

# Subset environmental samples for comparison
ESP_CTD_COI_envt_phyloseq <- prune_samples(sample_data(ESP_CTD_COI_phyloseq)$CTD_or_ESP %in% c("CTD","ESP"),ESP_CTD_COI_phyloseq)
ESP_CTD_COI_envt_phyloseq <- prune_taxa(taxa_sums(ESP_CTD_COI_envt_phyloseq) > 0, ESP_CTD_COI_envt_phyloseq)

ESP_CTD_COI_envt_asv_table <- as.data.frame(as(otu_table(ESP_CTD_COI_envt_phyloseq),"matrix"))
ESP_CTD_COI_envt_asv_table <- tibble::rownames_to_column(ESP_CTD_COI_envt_asv_table,"#OTUID")
write.table(ESP_CTD_COI_envt_asv_table, paste0(deicode_dir, "/ESP_CTD_COI_envt_asv_table_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
ESP_CTD_COI_envt_tax_table <- as.data.frame(as(tax_table(ESP_CTD_COI_envt_phyloseq),"matrix"))
ESP_CTD_COI_envt_tax_table <- tibble::rownames_to_column(ESP_CTD_COI_envt_tax_table,"#OTUID")
write.table(ESP_CTD_COI_envt_tax_table,paste0(deicode_dir, "/ESP_CTD_COI_envt_tax_table_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
ESP_CTD_COI_envtsamples_metadata <- as.data.frame(as.matrix(sample_data(ESP_CTD_COI_envt_phyloseq)))
ESP_CTD_COI_envtsamples_metadata <- tibble::rownames_to_column(ESP_CTD_COI_envtsamples_metadata,"#SampleID")
write.table(ESP_CTD_COI_envtsamples_metadata, paste0(deicode_dir, "/ESP_CTD_COI_envt_sample_data_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
```


### Run DEICODE in Qiime2

```{bash engine.opts = '-l'}
# Enter qiime2 environment
conda activate qiime2-2020.11

#Make biom file
biom convert -i ../DEICODE/COI/ESP_CTD_COI_envt_asv_table_for_biom.txt -o ../DEICODE/COI/table.from_txt_json.biom --table-type="OTU table" --to-json
#add metadata files to biom file
biom add-metadata -i ../DEICODE/COI/table.from_txt_json.biom -o ../DEICODE/COI/table.w_md.biom \
--observation-metadata-fp ../DEICODE/COI/ESP_CTD_COI_envt_tax_table_for_biom.txt \
--sample-metadata-fp ../DEICODE/COI/ESP_CTD_COI_envt_sample_data_for_biom.txt

#import into Qiime2
qiime tools import \
 --input-path ../DEICODE/COI/table.w_md.biom \
 --output-path ../DEICODE/COI/ESP_CTD_COI_master.biom.qza \
 --type FeatureTable[Frequency]

#run DEICODE
qiime deicode rpca \
    --i-table ../DEICODE/COI/ESP_CTD_COI_master.biom.qza \
    --p-n-components 3 \
    --p-min-feature-count 10 \
    --p-min-sample-count 500 \
    --o-biplot ../DEICODE/COI/ordination.qza \
    --o-distance-matrix ../DEICODE/COI/distance.qza
    
# Run PERMANOVA
 qiime diversity beta-group-significance \
    --i-distance-matrix ../DEICODE/COI/distance.qza \
    --m-metadata-file ../DEICODE/COI/ESP_CTD_COI_envt_sample_data_for_biom.txt \
    --m-metadata-column CTD_or_ESP \
    --p-method permanova \
    --o-visualization ../DEICODE/PERMANOVA_results/COI_ESP_CTD_PERMANOVA.qzv
```







### Plot COI DEICODE RPCA
```{r}
pco <- read_qza(here::here("DEICODE","COI","ordination.qza"))

label.PC1 <- paste0("PC1 (", round(pco$data$ProportionExplained$PC1, 3)*100,"%)")
label.PC1
label.PC2 <- paste0("PC2 (", round(pco$data$ProportionExplained$PC2, 3)*100,"%)")
label.PC2
label.PC3 <- paste0("PC3 (", round(pco$data$ProportionExplained$PC3, 3)*100,"%)")
label.PC3

## Prepare PCA data for ggplot

pca_metadata <- ESP_CTD_COI_envtsamples_metadata
pca_metadata %>% dplyr::rename(., "SampleID" = "#SampleID") -> pca_metadata
pca_data <- pco$data$Vectors
pca_data <- right_join(pca_data,pca_metadata,on = "SampleID")
pca_data <- subset(pca_data, !(is.na(pca_data$PC1)))

# Create a categorical depth
table(pca_data$depth)
pca_data %>%
  mutate(depth = as.numeric(depth)) %>%
  mutate(depth_cat = ifelse(depth <= 25, "0-25",
                            ifelse(depth <= 50, "25-50",
                                   ifelse(depth <= 201 & depth >=195, "195-200",
                                          "unknown")))) -> pca_data

# Shape guide for depths
depth_shapes <- c("0-25" = 21, "25-50" = 22, "195-200" = 24)
CTD_ESP_colors = c("CTD" = "#6a3d9a", "ESP" = "#33a02c")

ESP_CTD_COI_PCA <- ggplot(pca_data,aes(x=PC1,y=PC2,color = CTD_or_ESP,shape = depth_cat))+
  # Make fill for geom point be white if CN18F; if CN18S, then make it the color corresponding to ESP or CTD
  geom_point(size = 5, stroke = 2, fill = ifelse(pca_data$SAMPLING_cruise == "CN18F", "white", 
                                     ifelse(pca_data$CTD_or_ESP == "ESP", "#33a02c", "#6a3d9a")))+
  xlab(print(label.PC1))+
  ylab(print(label.PC2))+
  # annotate("text",  x=Inf, y = Inf, label = "(a)", vjust=1.5, hjust=1.3,size = 12)+
  scale_color_manual(values = CTD_ESP_colors)+
  scale_shape_manual(values = depth_shapes)+
  theme(panel.background = element_rect(fill = "white",size = 1),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.title = element_text(size = 20),
        axis.text = element_blank(),
        axis.ticks.length=unit(0.25, "cm"),
        axis.ticks=element_blank(),
        plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm"),
        axis.line.x.bottom = element_line(color = "black", size = 1),
        axis.line.y.left = element_line(color = "black", size = 1),
        legend.position = "none")

ESP_CTD_COI_PCA

ggsave(path = fig_dir, filename = "ESP_CTD_COI_PCA.png", ESP_CTD_COI_PCA,width = 7,height = 5)
```


# 18S

### Prep data for DEICODE
```{r}
# Set directory for saving DEICODE data
deicode_dir <- here::here("18S_12S_CO1_Analyses", "DEICODE", "18S")

# Subset environmental samples for comparison
ESP_CTD_18S_envt_phyloseq <- prune_samples(sample_data(ESP_CTD_18S_phyloseq)$CTD_or_ESP %in% c("CTD","ESP"),ESP_CTD_18S_phyloseq)
ESP_CTD_18S_envt_phyloseq <- prune_taxa(taxa_sums(ESP_CTD_18S_envt_phyloseq) > 0, ESP_CTD_18S_envt_phyloseq)

ESP_CTD_18S_envt_asv_table <- as.data.frame(as(otu_table(ESP_CTD_18S_envt_phyloseq),"matrix"))
ESP_CTD_18S_envt_asv_table <- tibble::rownames_to_column(ESP_CTD_18S_envt_asv_table,"#OTUID")
write.table(ESP_CTD_18S_envt_asv_table, paste0(deicode_dir, "/ESP_CTD_18S_envt_asv_table_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
ESP_CTD_18S_envt_tax_table <- as.data.frame(as(tax_table(ESP_CTD_18S_envt_phyloseq),"matrix"))
ESP_CTD_18S_envt_tax_table <- tibble::rownames_to_column(ESP_CTD_18S_envt_tax_table,"#OTUID")
write.table(ESP_CTD_18S_envt_tax_table, paste0(deicode_dir, "/ESP_CTD_18S_envt_tax_table_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
ESP_CTD_18S_envtsamples_metadata <- as.data.frame(as.matrix(sample_data(ESP_CTD_18S_envt_phyloseq)))
ESP_CTD_18S_envtsamples_metadata <- tibble::rownames_to_column(ESP_CTD_18S_envtsamples_metadata,"#SampleID")
write.table(ESP_CTD_18S_envtsamples_metadata, paste0(deicode_dir, "/ESP_CTD_18S_envt_sample_data_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
```

### Run DEICODE in Qiime2

```{bash engine.opts = '-l'}
# Enter qiime2 environment
conda activate qiime2-2020.11

#Make biom file
biom convert -i ../DEICODE/18S/ESP_CTD_18S_envt_asv_table_for_biom.txt -o ../DEICODE/18S/table.from_txt_json.biom --table-type="OTU table" --to-json
#add metadata files to biom file
biom add-metadata -i ../DEICODE/18S/table.from_txt_json.biom -o ../DEICODE/18S/table.w_md.biom \
--observation-metadata-fp ../DEICODE/18S/ESP_CTD_18S_envt_tax_table_for_biom.txt \
--sample-metadata-fp ../DEICODE/18S/ESP_CTD_18S_envt_sample_data_for_biom.txt

#import into Qiime2
qiime tools import \
--input-path ../DEICODE/18S/table.w_md.biom \
--output-path ../DEICODE/18S/ESP_CTD_18S_master.biom.qza \
--type FeatureTable[Frequency]

#run DEICODE
qiime deicode rpca \
--i-table ../DEICODE/18S/ESP_CTD_18S_master.biom.qza \
--p-n-components 3 \
--p-min-feature-count 10 \
--p-min-sample-count 500 \
--o-biplot ../DEICODE/18S/ordination.qza \
--o-distance-matrix ../DEICODE/18S/distance.qza

# Run PERMANOVA
 qiime diversity beta-group-significance \
    --i-distance-matrix ../DEICODE/18S/distance.qza \
    --m-metadata-file ../DEICODE/18S/ESP_CTD_18S_envt_sample_data_for_biom.txt \
    --m-metadata-column CTD_or_ESP \
    --p-method permanova \
    --o-visualization ../DEICODE/PERMANOVA_results/18S_ESP_CTD_PERMANOVA.qzv
```




### Plot 18S DEICODE RPCA
```{r}
# Set figure directory
fig_dir <- here::here("figures", "PCAs")

pco <- read_qza(here::here("DEICODE","18S","ordination.qza"))

label.PC1 <- paste0("PC1 (", round(pco$data$ProportionExplained$PC1, 3)*100,"%)")
label.PC1
label.PC2 <- paste0("PC2 (", round(pco$data$ProportionExplained$PC2, 3)*100,"%)")
label.PC2
label.PC3 <- paste0("PC3 (", round(pco$data$ProportionExplained$PC3, 3)*100,"%)")
label.PC3

## Prepare PCA data for ggplot

pca_metadata <- ESP_CTD_18S_envtsamples_metadata
pca_metadata %>% dplyr::rename(., "SampleID" = "#SampleID") -> pca_metadata
pca_data <- pco$data$Vectors
pca_data <- right_join(pca_data,pca_metadata,on = "SampleID")
pca_data <- subset(pca_data, !(is.na(pca_data$PC1)))

# Create a categorical depth
table(pca_data$depth)
pca_data %>%
  mutate(depth = as.numeric(depth)) %>%
  mutate(depth_cat = ifelse(depth <= 25, "0-25",
                            ifelse(depth <= 50, "25-50",
                                   ifelse(depth <= 201 & depth >=195, "195-200",
                                          "unknown")))) -> pca_data

# Shape guide for depths
depth_shapes <- c("0-25" = 21, "25-50" = 22, "195-200" = 24)
CTD_ESP_colors = c("CTD" = "#6a3d9a", "ESP" = "#33a02c")

ESP_CTD_18S_PCA <- ggplot(pca_data,aes(x=PC1,y=PC2,color = CTD_or_ESP,shape = depth_cat))+
  # Make fill for geom point be white if CN18F; if CN18S, then make it the color corresponding to ESP or CTD
  geom_point(size = 5, stroke = 2, fill = ifelse(pca_data$SAMPLING_cruise == "CN18F", "white", 
                                                 ifelse(pca_data$CTD_or_ESP == "ESP", "#33a02c", "#6a3d9a")))+
  xlab(print(label.PC1))+
  ylab(print(label.PC2))+
  # annotate("text",  x=Inf, y = Inf, label = "(a)", vjust=1.5, hjust=1.3,size = 12)+
  scale_color_manual(values = CTD_ESP_colors)+
  scale_shape_manual(values = depth_shapes)+
  theme(panel.background = element_rect(fill = "white",size = 1),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.title = element_text(size = 20),
        axis.text = element_blank(),
        axis.ticks.length=unit(0.25, "cm"),
        axis.ticks=element_blank(),
        plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm"),
        axis.line.x.bottom = element_line(color = "black", size = 1),
        axis.line.y.left = element_line(color = "black", size = 1),
        legend.position = "none")

ESP_CTD_18S_PCA

ggsave(path = fig_dir, filename = "ESP_CTD_18S_PCA.png", ESP_CTD_18S_PCA,width = 7,height = 5)
```


# 12S

### Prep data for DEICODE
```{r}
# Set directory for saving DEICODE data
deicode_dir <- here::here("18S_12S_CO1_Analyses", "DEICODE", "12S")

# Subset environmental samples for comparison
ESP_CTD_12S_envt_phyloseq <- prune_samples(sample_data(ESP_CTD_12S_phyloseq)$CTD_or_ESP %in% c("CTD","ESP"),ESP_CTD_12S_phyloseq)
ESP_CTD_12S_envt_phyloseq <- prune_taxa(taxa_sums(ESP_CTD_12S_envt_phyloseq) > 0, ESP_CTD_12S_envt_phyloseq)

ESP_CTD_12S_envt_asv_table <- as.data.frame(as(otu_table(ESP_CTD_12S_envt_phyloseq),"matrix"))
ESP_CTD_12S_envt_asv_table <- tibble::rownames_to_column(ESP_CTD_12S_envt_asv_table,"#OTUID")
write.table(ESP_CTD_12S_envt_asv_table, paste0(deicode_dir, "/ESP_CTD_12S_envt_asv_table_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
ESP_CTD_12S_envt_tax_table <- as.data.frame(as(tax_table(ESP_CTD_12S_envt_phyloseq),"matrix"))
ESP_CTD_12S_envt_tax_table <- tibble::rownames_to_column(ESP_CTD_12S_envt_tax_table,"#OTUID")
write.table(ESP_CTD_12S_envt_tax_table, paste0(deicode_dir, "/ESP_CTD_12S_envt_tax_table_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
ESP_CTD_12S_envtsamples_metadata <- as.data.frame(as.matrix(sample_data(ESP_CTD_12S_envt_phyloseq)))
ESP_CTD_12S_envtsamples_metadata <- tibble::rownames_to_column(ESP_CTD_12S_envtsamples_metadata,"#SampleID")
write.table(ESP_CTD_12S_envtsamples_metadata, paste0(deicode_dir, "/ESP_CTD_12S_envt_sample_data_for_biom.txt"),sep = "\t",row.names = FALSE,quote = FALSE)
```

### Run DEICODE in Qiime2

```{bash engine.opts = '-l'}
# Enter qiime2 environment
conda activate qiime2-2020.11

#Make biom file
biom convert -i ../DEICODE/12S/ESP_CTD_12S_envt_asv_table_for_biom.txt -o ../DEICODE/12S/table.from_txt_json.biom --table-type="OTU table" --to-json
#add metadata files to biom file
biom add-metadata -i ../DEICODE/12S/table.from_txt_json.biom -o ../DEICODE/12S/table.w_md.biom \
--observation-metadata-fp ../DEICODE/12S/ESP_CTD_12S_envt_tax_table_for_biom.txt \
--sample-metadata-fp ../DEICODE/12S/ESP_CTD_12S_envt_sample_data_for_biom.txt

#import into Qiime2
qiime tools import \
--input-path ../DEICODE/12S/table.w_md.biom \
--output-path ../DEICODE/12S/ESP_CTD_12S_master.biom.qza \
--type FeatureTable[Frequency]

#run DEICODE
qiime deicode rpca \
--i-table ../DEICODE/12S/ESP_CTD_12S_master.biom.qza \
--p-n-components 3 \
--p-min-feature-count 10 \
--p-min-sample-count 500 \
--o-biplot ../DEICODE/12S/ordination.qza \
--o-distance-matrix ../DEICODE/12S/distance.qza

# Run PERMANOVA
 qiime diversity beta-group-significance \
    --i-distance-matrix ../DEICODE/12S/distance.qza \
    --m-metadata-file ../DEICODE/12S/ESP_CTD_12S_envt_sample_data_for_biom.txt \
    --m-metadata-column CTD_or_ESP \
    --p-method permanova \
    --o-visualization ../DEICODE/PERMANOVA_results/12S_ESP_CTD_PERMANOVA.qzv
```




### Plot 12S DEICODE RPCA
```{r}
# Set figure directory
fig_dir <- here::here("figures", "PCAs")

pco <- read_qza(here::here("DEICODE","12S","ordination.qza"))

label.PC1 <- paste0("PC1 (", round(pco$data$ProportionExplained$PC1, 3)*100,"%)")
label.PC1
label.PC2 <- paste0("PC2 (", round(pco$data$ProportionExplained$PC2, 3)*100,"%)")
label.PC2
label.PC3 <- paste0("PC3 (", round(pco$data$ProportionExplained$PC3, 3)*100,"%)")
label.PC3

## Prepare PCA data for ggplot

pca_metadata <- ESP_CTD_12S_envtsamples_metadata
pca_metadata %>% dplyr::rename(., "SampleID" = "#SampleID") -> pca_metadata
pca_data <- pco$data$Vectors
pca_data <- right_join(pca_data,pca_metadata,on = "SampleID")
pca_data <- subset(pca_data, !(is.na(pca_data$PC1)))

# Create a categorical depth
table(pca_data$depth)
pca_data %>%
  mutate(depth = as.numeric(depth)) %>%
  mutate(depth_cat = ifelse(depth <= 25, "0-25",
                            ifelse(depth <= 50, "25-50",
                                   ifelse(depth <= 201 & depth >=195, "195-200",
                                          "unknown")))) -> pca_data

# Shape guide for depths
depth_shapes <- c("0-25" = 21, "25-50" = 22, "195-200" = 24)
CTD_ESP_colors = c("CTD" = "#6a3d9a", "ESP" = "#33a02c")

ESP_CTD_12S_PCA <- ggplot(pca_data,aes(x=PC1,y=PC2,color = CTD_or_ESP,shape = depth_cat))+
  # Make fill for geom point be white if CN18F; if CN12S, then make it the color corresponding to ESP or CTD
  geom_point(size = 5, stroke = 2, fill = ifelse(pca_data$SAMPLING_cruise == "CN18F", "white", 
                                                 ifelse(pca_data$CTD_or_ESP == "ESP", "#33a02c", "#6a3d9a")))+
  xlab(print(label.PC1))+
  ylab(print(label.PC2))+
  # annotate("text",  x=Inf, y = Inf, label = "(a)", vjust=1.5, hjust=1.3,size = 12)+
  scale_color_manual(values = CTD_ESP_colors)+
  scale_shape_manual(values = depth_shapes)+
  theme(panel.background = element_rect(fill = "white",size = 1),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.title = element_text(size = 20),
        axis.text = element_blank(),
        axis.ticks.length=unit(0.25, "cm"),
        axis.ticks=element_blank(),
        plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm"),
        axis.line.x.bottom = element_line(color = "black", size = 1),
        axis.line.y.left = element_line(color = "black", size = 1),
        legend.position = "none")

ESP_CTD_12S_PCA

ggsave(path = fig_dir, filename = "ESP_CTD_12S_PCA.png", ESP_CTD_12S_PCA,width = 7,height = 5)
```
# Create a legend manually
```{r}
PCA_legend <- ggplot(pca_data) +
  # ESP vs. CTD
  annotate("text",label = "Sampling Method", x = 0.9, y = 2.8,size = 10,adj = 0)+ # Title ESP or CTD
  annotate("point", x = 1, y = 1, shape = 21, colour = "#33a02c", fill = "#33a02c", size = 7, stroke = 3)+ # ESP
  annotate("point", x = 1, y = 2, shape = 21, colour = "#6a3d9a", fill = "#6a3d9a", size = 7, stroke = 3)+ # CTD
  annotate("text",label = "Automated (ESP)", x = 1.2, y = 2,size = 7.5,adj = 0)+ #ESP
  annotate("text",label = "CTD/Niskin Rosette", x = 1.2, y = 1,size = 7.5,adj = 0)+ #CTD
  # Spring vs. Fall
  annotate("text",label = "Season", x = 4.9, y = 2.8,size = 10,adj = 0)+ # Title seasonality
  annotate("point", x = 5, y = 2, shape = 21, colour = "black", fill = "black", size = 7, stroke = 3)+ # Spring
  annotate("point", x = 5, y = 1, shape = 21, colour = "black", fill = "white", size = 7, stroke = 3)+ # Fall
  annotate("text",label = "Spring", x = 5.2, y = 2,size = 7.5,adj = 0)+ #Spring
  annotate("text",label = "Fall", x = 5.2, y = 1,size = 7.5,adj = 0)+ #Fall
  # Depths depth_shapes <- c("0-25" = 21, "25-50" = 22, "195-200" = 24)
  annotate("text",label = "Depth", x = 7.9, y = 2.8,size = 10,adj = 0)+ # Title depths
  annotate("point", x = 8, y = 2, shape = 21, colour = "black", fill = "black", size = 7, stroke = 3)+ # 0-25 m
  annotate("text",label = "0-25 m", x = 8.2, y = 2,size = 7.5,adj = 0)+ #0-25 m
  annotate("point", x = 8, y = 1.2, shape = 22, colour = "black", fill = "black", size = 7, stroke = 3)+ # 25-50 m
  annotate("text",label = "25-50 m", x = 8.2, y = 1.2,size = 7.5,adj = 0)+ #25-50 m
  annotate("point", x = 8, y = 0.4, shape = 24, colour = "black", fill = "black", size = 7, stroke = 3)+ # 195-200 m
  annotate("text",label = "195-200 m", x = 8.2, y = 0.4,size = 7.5,adj = 0)+ #195-200 m
  
  xlim(0,10)+
  ylim(0.2,3.2)+
    theme(panel.background = element_rect(fill="white"),
        panel.border = element_rect(colour = "black", fill=NA, size=2),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(), 
        legend.position = "none",
        plot.margin = unit(c(0,0,0,0),"pt"))
```

# Combine all PCA figures
```{r}
combined_PCA_gg <- annotate_figure(ggarrange(ggarrange(ESP_CTD_12S_PCA, ESP_CTD_18S_PCA, 
                                                 ESP_CTD_COI_PCA, ESP_CTD_12S_PCA, ncol = 2, nrow = 2,
                                                 labels = c("(a)", "(b)", "(c)", "(d)"), label.x = 0.85, label.y = 0.97,
                                                 font.label = list(size = 35, color = "black", face = "bold")),
                                                 PCA_legend, nrow = 2, ncol = 1, heights = c(7,1))
                                   )+
  annotate("text", x = 0.25, y = 0.8, size = 15, angle = 45, color = "red", label = "PLACEHOLDER FOR 16S")


ggsave(combined_PCA_gg, height = 16, width = 16, path = fig_dir, filename = "combined_PCA_fig.png", device = "png")

```




